---
title: "Getting Started with istatlab"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with istatlab}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE  # Set to TRUE when package is fully functional
)
```

```{r setup}
library(istatlab)
```

# Introduction

The **istatlab** package provides a comprehensive toolkit for working with Italian labour market data from ISTAT (Istituto Nazionale di Statistica). This vignette demonstrates the main functionality of the package.

## Checking API Connectivity

Before starting, it's good practice to check if the ISTAT API is accessible:

```{r}
# Check if ISTAT API is available
if (check_istat_api()) {
  message("ISTAT API is accessible!")
} else {
  message("ISTAT API is not accessible. Check your internet connection.")
}
```

## Downloading Metadata

The first step is typically to download metadata to understand what datasets are available:

```{r}
# Download metadata for available datasets
metadata <- download_metadata()

# View the structure
head(metadata)

# Look for employment-related datasets
employment_datasets <- metadata[grepl("employment|lavoro", Name.it, ignore.case = TRUE)]
print(employment_datasets[, .(id, Name.it)])
```

## Downloading Data

Once you've identified the datasets you need, you can download the actual data:

```{r}
# Download data for a specific dataset
# Using dataset "150_915" as an example (quarterly employment statistics)
data <- download_istat_data("150_915", start_time = "2020")

# View the raw data structure
str(data)
head(data)
```

## Processing and Labeling Data

Raw ISTAT data contains codes that need to be translated to meaningful labels:

```{r}
# Apply labels to make data more readable
processed_data <- apply_labels(data)

# View the processed data
str(processed_data)
head(processed_data)
```

## Basic Analysis

The package provides various analysis functions:

### Summary Statistics

```{r}
# Calculate summary statistics
summary_stats <- calculate_summary_stats(processed_data)
print(summary_stats)
```

### Growth Rate Analysis

```{r}
# Calculate growth rates
data_with_growth <- calculate_growth_rates(processed_data, type = "annual")

# View growth rates
head(data_with_growth[, .(tempo, valore, growth_annual)])
```

### Trend Analysis

```{r}
# Analyze trends
trend_results <- analyze_trends(processed_data, method = "linear")
print(trend_results)
```

## Forecasting

Generate forecasts for future periods:

```{r}
# Generate 12-period forecast using auto.arima
forecast_results <- forecast_series(processed_data, 
                                   periods = 12, 
                                   method = "auto.arima")

# View forecast summary
forecast_summary <- create_forecast_summary(forecast_results)
head(forecast_summary)
```

## Visualization

Create publication-ready plots:

```{r}
# Basic time series plot
ts_plot <- create_time_series_plot(processed_data,
                                  title = "Italian Labour Market Data",
                                  subtitle = "Source: ISTAT",
                                  y_label = "Value")

print(ts_plot)
```

```{r}
# Forecast plot
forecast_plot <- create_forecast_plot(processed_data, forecast_results,
                                     title = "Labour Market Forecast")

print(forecast_plot)
```

```{r}
# Growth rate plot
growth_plot <- create_growth_plot(data_with_growth, 
                                 growth_col = "growth_annual",
                                 title = "Annual Growth Rates")

print(growth_plot)
```

## Working with Multiple Datasets

You can also download and analyze multiple datasets:

```{r}
# Download multiple datasets
dataset_ids <- c("150_915", "151_914")
multiple_data <- download_multiple_datasets(dataset_ids, start_time = "2020")

# Process each dataset
processed_multiple <- lapply(multiple_data, apply_labels)

# Combine for comparison (if they have compatible structure)
# This would depend on the specific datasets and their dimensions
```

## Advanced Features

### Dashboard Creation

```{r}
# Create a comprehensive dashboard
dashboard <- create_dashboard_plot(processed_data,
                                  panels = c("timeseries", "growth"),
                                  title = "Labour Market Dashboard")

print(dashboard)
```

### Structural Break Detection

```{r}
# Detect structural breaks in the time series
breaks <- detect_structural_breaks(processed_data)
print(breaks)
```

## Best Practices

1. **Always check API connectivity** before attempting downloads
2. **Cache metadata locally** - the package does this automatically
3. **Validate data** after processing using `validate_istat_data()`
4. **Filter data by time** to focus on relevant periods
5. **Use appropriate forecasting methods** based on your data characteristics
6. **Create reproducible workflows** by saving dataset IDs and parameters

## Troubleshooting

Common issues and solutions:

- **Connection timeouts**: Increase timeout values or check internet connection
- **Memory issues**: Process data in chunks for very large datasets
- **Missing labels**: Ensure codelists are downloaded and cached

## Next Steps

- Explore the full range of available datasets in ISTAT's catalogue
- Experiment with different forecasting methods
- Customize visualizations for your specific needs
- Contribute to the package development on GitHub

For more information, see the package documentation and examples.